---
title: "Global_QC"
author: "Rasmus Jensen, Anne Cathrine Baun Thuesen, Sara Stinson, Yun Huang, Helene Bæk Juel, Evelina Stankevic"
date: "29/03/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Load packages
source(here::here("R/package-loading.R"))
```

## Load full dataset
```{r load-data, include=FALSE}
# Load data from raw NPX data
holbaek_serum <- OlinkAnalyze::read_NPX(here::here("data-raw/20210038_Stinson_NPX.xlsx"))

# Add cohort column
holbaek <- holbaek_serum %>%
    dplyr::mutate(cohort = "HOL") %>% #Here we add the column cohort and name them all "HOL"
    dplyr::mutate(cohort = case_when(str_detect(SampleID, "CONTROL") ~ "Control", TRUE ~ "HOL")) #Here we name all the control samples "control" instead of "HOL"

#Loading microbliver bridged samples 
micro <- readRDS(here::here("data-raw/20202249_20212246_NPX_Bridge_Normalized.rds")) %>%
  dplyr::select(-Project, -Adj_factor)
micro_b3 <- readRDS(here::here("data-raw/Q-00557_NPX_Bridge_Normalized.rds"))%>%
  dplyr::select(-Project, -Adj_factor)

# replace typo TIEP80EC --> TIPS80EC
micro <- micro %>%
    mutate(SampleID = sub("TIEP80EC", "TIPS80EC", SampleID))
# replace typo TIPSKC/93 --> TIPS93
micro <- micro %>%
    mutate(SampleID = sub("TIPSKC/93", "TIPS93", SampleID))

# Add cohort column
micro <- micro %>%
    mutate(cohort = sub("[0-9]{1,4}.*$", "", SampleID)) %>% # substitute 1-4 numbers and everything after this (.*$) with nothing
    mutate(cohort = sub("_.*$", "", cohort)) # substitute anything after a _ with nothing to remove long control names and just keep "CONTROL"

micro_b3 <- micro_b3 %>%
    mutate(cohort = sub("[0-9]{1,4}.*$", "", SampleID)) %>% # substitute 1-4 numbers and everything after this (.*$) with nothing
    mutate(cohort = sub("_.*$", "", cohort)) # substitute anything after a _ with nothing to remove long control names and just keep "CONTROL"

#merge the three dataframes
merged_olink <- rbind(micro, micro_b3, holbaek)

# change cytokine names
# first remove all -
# then change spaces to _
# then rename alpha/beta/gamma to a/b/g
merged_olink <- merged_olink %>%
    mutate(Assay = gsub("-", "", Assay)) %>%
    mutate(Assay = gsub(" ", "_", Assay)) %>%
    mutate(Assay = gsub("alpha", "a", Assay)) %>%
    mutate(Assay = gsub("beta", "b", Assay)) %>%
    mutate(Assay = gsub("gamma", "g", Assay))

# change values < LOD of that assay to 50% LOD
merged_olink <- merged_olink %>%
    mutate(corrected_NPX = if_else(LOD > NPX, LOD/2, NPX)) %>% #include this to df but not use this for QC
  dplyr::mutate(Panel = replace(Panel, Panel == "Olink Target 96 Inflammation", "Olink Inflammation"))

```

# Wrangling and QC

## Visualize potential outliers by IQR vs. sample median
```{r QC_plot}
olink_qc_plot(merged_olink)
```

Some samples have a QC warning (red) and several appear to be outliers, but no samples are both outliers and have QC warning.

We can see that out of only 37 samples with QC warning, some ppts with multiple samples have QC warning in all or several samples, presumably due to some underlying disturbance in their blood sample. We suggest to exclude all samples from ppts with more than one QC warning, however we do not exclude any samples ourselves.



```{r}

# add a column for sample type (Lihep EDTA etc, check the last sample manifest)
merged_olink2<-merged_olink %>% 
    dplyr::mutate(cohort = replace(cohort, cohort == "Control", "CONTROL")) %>% 
    dplyr::mutate(Sample_type = case_when(
        cohort %in% c("AHEP", "Bridge", "HCOT", "MLGB","PK", "PRF", "RDC", "SIP", "TIPS", "ProD", "Bridge", "CONTROL", "HOL-F", "HOL-N") ~ "Plasma_EDTA", 
        cohort %in% c("ALD", "ALCO", "HP", "RFX") ~ "Plasma_LiHep", 
        cohort %in% c("HOL") ~ "Serum"
    ))

#This code chunk will create a global missing frequency variable
merged_olink3 <- merged_olink2 %>%
    group_by(Assay, Sample_type) %>% 
  dplyr::mutate(no_missing_obs = sum(is.na(NPX) | NPX < LOD),
         total_obs = length(NPX)) %>%
  dplyr::mutate(MissingProp = no_missing_obs/total_obs) # calculates the global MissingFreq


#Writing csv file to be split in the next script
write.csv(merged_olink3, here::here("data/20220330Global_QC_olink_wBatch3.csv"), row.names = F)
```


#Making dataset order by assays
```{r}
## Creating a panel overview of the proteins included in the olink dataset
olink_proteins <- merged_olink3 %>% # using the df in the long format
  dplyr::group_by(Assay) %>%
  dplyr::select(UniProt, OlinkID, LOD, Panel, Panel_Version, MissingProp) %>% #These I think would be important to include
  dplyr::distinct(Assay, .keep_all = TRUE) %>% # This keeps one value for each protein in the long format
  arrange(Assay) # Orders proteins

## Save metadata
write.table(olink_proteins, file = here::here("data/20220330_olink_proteins_wBatch3.txt"), sep = "\t", quote = F)
```
