---
title: "MicrobLiver_Olink_QC_Bridge"
author: "Sara Elizabeth Stinson"
date: "19/10/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source(here::here("R/Project_Files/package-loading.R"))
```

```{r LOAD RAW DATA, echo = FALSE, warning = FALSE}

library(OlinkAnalyze)

## LOAD MicrobLiver BATCH 1 (N = 1,808)
MLB1_INF <- OlinkAnalyze::read_NPX(filename = "/Volumes/SUN-CBMR-Hansen-Group/Projects/MicrobLiver/Results/Olink/Raw_data_reports/20202249_Juel_NPX.xlsx")

## LOAD MicrobLiver BATCH 2 (N = 688)
MLB2 <- OlinkAnalyze::read_NPX(filename = "/Volumes/SUN-CBMR-Hansen-Group/Projects/MicrobLiver/Results/Olink/Raw_data_reports/20212246_Juel_NPX.xlsx")

```

Bridging normalization: One of the dataframes is adjusted to another using overlapping samples (bridge samples). The overlapping samples need to be named the same between the dataframes and adjustment is made using the median of the paired differences between the bridge samples in the two data frames. The two dataframes are inputs df1 and df2, the one being adjusted to is specified in the input reference_project and the overlapping samples are specified in overlapping_samples_df1. Only overlapping_samples_df1 should be input, no matter which dataframe is used as reference_project. 

```{r BRIDGE BATCHES, echo = FALSE, warning = FALSE}

## SELECT CVD PANEL
MLB2_CVD <- MLB2 %>%
  dplyr::filter(Panel == 'Olink Cardiovascular II')

## SELECT INFLAMMATION PANEL
MLB2_INF <- MLB2 %>%
  dplyr::filter(Panel == 'Olink Inflammation')

## ADD PROJECT NUMBER
MLB1_INF_df1 <- MLB1_INF %>% dplyr::mutate(Project = '20202249')
MLB2_INF_df2 <- MLB2_INF %>% dplyr::mutate(Project = '20212246')

## RENAME BRIDGE SAMPLES (CONSISTENCY)
MLB2_INF_df2r <- MLB2_INF_df2 %>%
  dplyr::mutate(SampleID = gsub("BRIDGE0", "Bridge_", SampleID)) %>%
  dplyr::mutate(SampleID = gsub("BRIDGE", "Bridge_", SampleID))

## IDENTIFY OVERLAP OF 16 BRIDGE SAMPLES 
overlap_samples <- intersect((MLB1_INF_df1 %>% filter(!grepl("control", SampleID, ignore.case=T)))$SampleID,
                             (MLB2_INF_df2r %>% filter(!grepl("control", SampleID, ignore.case=T)))$SampleID)
# Find overlap in bridge samples, but exclude Olink controls

## BRIDGING NORMALIZATION
bridged_df <- OlinkAnalyze::olink_normalization(df1 = MLB1_INF_df1,
                      df2 = MLB2_INF_df2r,
                      overlapping_samples_df1 = overlap_samples,
                      df1_project_nr = '20202249',
                      df2_project_nr = '20212246',
                      reference_project = '20202249')

# The second batch will be adjusted too the first batch, according to "reference_project"

## SAVE AS .RDS
saveRDS(object = bridged_df, file =
          "/Volumes/SUN-CBMR-Hansen-Group/Projects/MicrobLiver/Results/Olink/Bridge_Normalized/20202249_20212246_NPX_Bridge_Normalized.rds")

## LOADING FOR QC
ML_B1_B2 <- readRDS(file = "/Volumes/SUN-CBMR-Hansen-Group/Projects/MicrobLiver/Results/Olink/Bridge_Normalized/20202249_20212246_NPX_Bridge_Normalized.rds")

```

